---
title: "Introduction to LLM_EnrichViz"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to LLM_EnrichViz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE only when API keys are available
)
```

# Overview

**LLM_EnrichViz** is an R package that bridges quantitative gene set enrichment
statistics with the interpretive power of Large Language Models (LLMs).  It
runs four complementary enrichment methods on your transcriptomic data and
dispatches each result to an independent LLM agent for biological interpretation.
A final synthesising LLM call integrates all agent outputs into a unified
mechanistic model, a confidence-ranked pathway list, and an interactive network
visualisation.

## Key features

| Feature | Description |
|---------|-------------|
| **Four enrichment methods** | ORA, GSEA, GSVA, ROAST |
| **Multiple pathway databases** | KEGG, GO, Reactome, WikiPathways |
| **Four LLM providers** | Gemini, OpenAI (GPT), Anthropic (Claude), DeepSeek |
| **Multi-agent parallelism** | All (method × database) pairs run simultaneously |
| **Cross-method integration** | Confidence tiers based on method convergence |
| **Interactive visualisation** | `visNetwork` HTML network |
| **Automated report** | Self-contained HTML via RMarkdown |

---

# Installation

## From GitHub

```{r install-github}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("yourname/LLM_EnrichViz")
```

## Bioconductor dependencies

```{r install-bioc}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "clusterProfiler",
  "ReactomePA",
  "GSVA",
  "limma",
  "org.Hs.eg.db",   # human
  "org.Mm.eg.db"    # mouse
))
```

## CRAN dependencies

```{r install-cran}
install.packages(c(
  "msigdbr", "httr", "R6", "future", "furrr",
  "progressr", "dplyr", "tidyr", "ggplot2",
  "visNetwork", "stringr", "rmarkdown", "tibble"
))
```

---

# Quick-start examples

## Example 1 — ORA only (simplest case)

This example requires only a list of differentially expressed genes.

```{r example-ora}
library(LLM_EnrichViz)

# A short list of human DEGs (gene symbols)
deg_list <- c(
  "TP53", "MYC", "BRCA1", "EGFR", "VEGFA", "CDK2", "MDM2",
  "BCL2", "PTEN", "RB1", "AKT1", "STAT3", "NFkB1", "CCND1"
)

LLM_EnrichViz(
  deg_list            = deg_list,
  gene_type           = "SYMBOL",
  organism            = "human",
  analyses            = "ORA",
  databases           = c("KEGG", "GO"),
  pvalue              = 0.05,
  ont                 = "BP",
  llm_provider        = "openai",
  llm_model           = "gpt-4o",
  api_key_file        = "~/.openai_key",     # one line: your API key
  experimental_design = "Breast cancer cell line vs normal, bulk RNA-Seq",
  output_dir          = "ora_results"
)
```

Results are written to `ora_results/`:

```
ora_results/
├── ORA_KEGG_results.txt
├── ORA_GO_results.txt
├── ORA_KEGG_barplot.pdf
├── ORA_GO_barplot.pdf
├── agent_responses.txt
├── final_response.txt
├── integrated_network.html
└── report_template.html
```

## Example 2 — GSEA with a ranked gene list

GSEA requires all genes ranked by a continuous statistic (Wald statistic,
log2FC, signed -log10 p-value, etc.).

```{r example-gsea}
library(LLM_EnrichViz)

# Build a ranked list from DESeq2 results
# (assuming `deseq_res` is a DESeqResults object)
ranked_list <- deseq_res$stat               # Wald statistics
names(ranked_list) <- rownames(deseq_res)   # SYMBOL or ENSEMBL
ranked_list <- sort(ranked_list, decreasing = TRUE)
# Remove NA values
ranked_list <- ranked_list[!is.na(ranked_list)]

LLM_EnrichViz(
  ranked_list         = ranked_list,
  gene_type           = "SYMBOL",
  organism            = "human",
  analyses            = "GSEA",
  databases           = c("KEGG", "Reactome"),
  pvalue              = 0.05,
  llm_provider        = "claude",
  llm_model           = "claude-sonnet-4-6",
  api_key_file        = "~/.claude_key",
  experimental_design = "PBMC: lupus vs healthy, bulk RNA-Seq, n=50",
  output_dir          = "gsea_results"
)
```

## Example 3 — GSVA + ROAST (matrix-based methods)

GSVA and ROAST both require the full expression matrix.

```{r example-gsva-roast}
library(LLM_EnrichViz)
library(limma)

# expr_matrix : normalised log2 CPM matrix (genes x samples)
# sample_info : data frame with a 'condition' column ("case" / "control")

design <- model.matrix(~ condition, data = sample_info)
contrast <- c(0, 1)   # test 'conditioncase'

LLM_EnrichViz(
  expression_data     = expr_matrix,
  deg_list            = sig_genes,       # for ORA as well
  ranked_list         = rank_vec,        # for GSEA as well
  gene_type           = "SYMBOL",
  organism            = "human",
  design_matrix       = design,
  contrast_vector     = contrast,
  analyses            = c("ORA", "GSEA", "GSVA", "ROAST"),
  databases           = c("KEGG", "GO", "Reactome"),
  pvalue              = 0.05,
  llm_provider        = "gemini",
  llm_model           = "gemini-1.5-flash-latest",
  api_key_file        = "~/.gemini_key",
  experimental_design = "Uveitis vs healthy controls, PBMC, bulk RNA-Seq",
  output_dir          = "full_analysis"
)
```

## Example 4 — Using DeepSeek

```{r example-deepseek}
LLM_EnrichViz(
  deg_list     = deg_list,
  gene_type    = "SYMBOL",
  organism     = "mouse",
  analyses     = c("ORA", "GSEA"),
  databases    = c("KEGG", "GO"),
  llm_provider = "deepseek",
  llm_model    = "deepseek-chat",
  api_key_file = "~/.deepseek_key",
  output_dir   = "mouse_analysis"
)
```

---

# Method details

## ORA — Over-Representation Analysis

ORA applies a **hypergeometric (Fisher's exact) test** to ask: given a
background gene universe, are members of a gene set over-represented among
the DEGs?

**Strengths:** simple, fast, widely used, easy to interpret.  
**Limitations:** discards expression magnitude; sensitive to the significance
threshold used to define DEGs; gene sets with many members have an inherent
advantage.

**Required input:** `deg_list` — a character vector of significant gene IDs.

**Key output columns:**

| Column | Meaning |
|--------|---------|
| `Description` | Gene set name |
| `pvalue` | Hypergeometric p-value |
| `qvalue` | BH-adjusted p-value |
| `Count` | Overlap size |
| `Gene` | Overlapping gene symbols |

## GSEA — Gene Set Enrichment Analysis

GSEA uses the **full ranked gene list** to compute a running enrichment
score (ES) by walking from the highest- to lowest-ranked gene and noting
when set members are encountered.  The ES peak is the enrichment score;
statistical significance is assessed by permutation (or by the preranked
approach here).

**Strengths:** uses the entire dataset; captures subtle coordinated shifts
that ORA misses; directional (NES > 0 = activated, NES < 0 = repressed).  
**Limitations:** sensitive to ranking metric choice; slow for very large
databases.

**Required input:** `ranked_list` — a **named, sorted** numeric vector.

**Key output columns:**

| Column | Meaning |
|--------|---------|
| `NES` | Normalised enrichment score |
| `enrichmentScore` | Raw ES |
| `pvalue` | Permutation p-value |
| `Gene` | Leading-edge (core enrichment) genes |

## GSVA — Gene Set Variation Analysis

GSVA transforms the expression matrix into a **gene set score matrix**
(gene sets × samples), enabling pathway-level differential analysis and the
study of **inter-sample heterogeneity** in pathway activity.

**Strengths:** sample-level resolution; detects sub-population differences;
compatible with standard differential expression frameworks (limma/edgeR).  
**Limitations:** scores are relative and not directly comparable across
different datasets; computationally heavier than ORA/GSEA.

**Required input:** `expression_data` — a numeric matrix (genes × samples).

**Key output columns:**

| Column | Meaning |
|--------|---------|
| `GeneSet` | Gene set name |
| `MeanScore` | Mean GSVA score across samples |
| `SDScore` | Standard deviation (inter-sample variability) |
| `Genes` | Genes in the set present in expression data |

## ROAST — Rotation-based Gene Set Testing

ROAST uses **random rotations of residuals** to test whether genes in a
set show coordinated differential expression beyond what is expected by
chance.  Unlike competitive methods (ORA, GSEA), ROAST is a **self-contained
test** — it does not compare to a background.

**Strengths:** accounts for intra-set gene correlations; directional
(Up/Down/Mixed); robust to small sample sizes.  
**Limitations:** requires a design matrix and contrast; computationally
intensive (1 000 rotations by default).

**Required inputs:** `expression_data`, `design_matrix`, `contrast_vector`.

**Key output columns:**

| Column | Meaning |
|--------|---------|
| `GeneSet` | Gene set name |
| `PValue` | Rotation-based p-value |
| `Direction` | Up / Down / Mixed |
| `Genes` | Member genes found in expression data |

---

# LLM Providers

## Setting up API keys

Each provider requires its own API key, stored in a plain-text file
(one key per line; only the first line is read):

```bash
# Gemini
echo "AIza..." > ~/.gemini_key

# OpenAI
echo "sk-..." > ~/.openai_key

# Claude (Anthropic)
echo "sk-ant-..." > ~/.claude_key

# DeepSeek
echo "sk-..." > ~/.deepseek_key
```

## Provider comparison

| Provider | Speed | Cost | Strengths |
|----------|-------|------|-----------|
| **Gemini 1.5 Flash** | Fast | Low | Large context window, fast iteration |
| **GPT-4o** | Medium | Medium | Strong reasoning, reliable JSON |
| **Claude Sonnet** | Fast | Medium | Long context, nuanced interpretation |
| **DeepSeek Chat** | Fast | Very low | Cost-effective, scientific reasoning |

## Changing the model

```{r model-select}
# Use a more capable model for the final synthesis step
LLM_EnrichViz(
  ...,
  llm_provider = "openai",
  llm_model    = "gpt-4o",   # or "gpt-4-turbo", "o1-preview", etc.
  temperature  = 0           # 0 = deterministic, 1 = creative
)
```

---

# Output files

After a complete run, `output_dir` contains:

| File | Description |
|------|-------------|
| `ORA_{DB}_results.txt` | ORA filtered results per database |
| `GSEA_{DB}_results.txt` | GSEA filtered results per database |
| `GSVA_{DB}_summary.txt` | GSVA score summary per database |
| `GSVA_{DB}_scores_matrix.txt` | Full GSVA score matrix |
| `ROAST_{DB}_results.txt` | ROAST results per database |
| `ORA_{DB}_barplot.pdf` | ORA bar chart |
| `GSEA_{DB}_bubble.pdf` | GSEA bubble plot |
| `GSVA_{DB}_heatmap.pdf` | GSVA heatmap |
| `ROAST_{DB}_barplot.pdf` | ROAST bar chart |
| `agent_responses.txt` | Raw LLM agent responses |
| `final_response.txt` | Integrated LLM synthesis |
| `integrated_network.html` | Interactive visNetwork graph |
| `analysis_parameters.txt` | Run configuration |
| `report_template.html` | Self-contained HTML report |

---

# Interpreting the integrated network

The interactive network (`integrated_network.html`) is built from the TSV
block in the LLM's final integration response.  Each node is coloured by type:

- **Orange nodes** — genes (DEGs or hub genes)
- **Blue nodes** — pathways, GO terms, or processes

Edge labels indicate the relationship type: *activates*, *represses*,
*associated with*, *targets*.

**Confidence tiers** (shown in edge tooltips):
- **Tier 1** (highest): gene set detected by 4 methods
- **Tier 2**: detected by 3 methods
- **Tier 3**: detected by 2 methods
- **Tier 4** (lowest): detected by 1 method

---

# Frequently asked questions

**Q: Can I run a single method without providing data for others?**  
A: Yes.  Set `analyses = "ORA"` (or any subset) and only provide the required
inputs for that method.  Missing data for other methods generates a warning
and the method is skipped.

**Q: What gene identifier types are supported?**  
A: `SYMBOL`, `ENSEMBL`, and `ENTREZID`.  ENSEMBL version suffixes
(e.g., `ENSG00000139618.12`) are stripped automatically.

**Q: My GSEA ranked list has ENSEMBL IDs, not symbols.  Is that OK?**  
A: Yes — set `gene_type = "ENSEMBL"` and LLM_EnrichViz will map all inputs
to ENTREZID/SYMBOL internally.

**Q: The LLM returns an error (503 / rate limit). What happens?**  
A: Each agent automatically retries up to 3 times with a doubled sleep
interval.  After max retries, the agent records a failure and the pipeline
continues with the remaining agents.

**Q: Can I use a custom pathway database?**  
A: Not directly through the main function yet.  The `get_gene_sets()`
function can be extended to include custom MSigDB collections or user-defined
gene set lists.

**Q: How long does a full run take?**  
A: A typical run with 4 methods × 3 databases = 12 agents takes 5–15 minutes
depending on LLM response times and whether all analyses are enabled.

---

# Session information

```{r session-info, eval=TRUE}
sessionInfo()
```
