---
title: "LLM_EnrichViz: LLM-Powered Multi-Method Gene Set Enrichment Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{LLM_EnrichViz: LLM-Powered Multi-Method Gene Set Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # set to TRUE only in interactive use (requires API key)
)
```

# Introduction

**LLM_EnrichViz** is an R package that integrates four complementary gene set
enrichment methods — ORA, GSEA, GSVA, and ROAST — with Large Language Models
(LLMs). Each analysis is independently performed and then interpreted by a
dedicated LLM agent. A final orchestrator LLM synthesises all agent responses
into a unified mechanistic model with cross-method confidence levels, an
interactive network, and an HTML report.

## Why four methods?

| Method | Input | Key strength |
|--------|-------|-------------|
| **ORA** | DEG list | Fast; interpretable threshold-based test |
| **GSEA** | Ranked gene list | Captures sub-threshold coordinated signals; NES gives direction |
| **GSVA** | Full expression matrix | Sample-level pathway scores; reveals cohort heterogeneity |
| **ROAST** | Full expression matrix + design | Directional coordination; robust to within-set gene correlation |

Using all four methods together increases biological confidence: a pathway
supported by every method is far more reliable than one detected by a single
approach.

## Supported LLM providers

| Provider | Models | Environment variable |
|----------|--------|----------------------|
| Google Gemini | `gemini-1.5-flash-latest`, `gemini-1.5-pro-latest` | `GEMINI_API_KEY` |
| OpenAI | `gpt-4o`, `gpt-4-turbo`, `gpt-3.5-turbo` | `OPENAI_API_KEY` |
| Anthropic Claude | `claude-opus-4-6`, `claude-sonnet-4-6`, `claude-haiku-4-5-20251001` | `ANTHROPIC_API_KEY` |
| DeepSeek | `deepseek-chat`, `deepseek-reasoner` | `DEEPSEEK_API_KEY` |

---

# Installation

## From GitHub (development version)

```{r install-github}
if (!requireNamespace("remotes", quietly = TRUE))
  install.packages("remotes")

remotes::install_github("yourname/LLM_EnrichViz")
```

## Bioconductor dependencies

```{r install-bioc}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "clusterProfiler",
  "ReactomePA",
  "GSVA",
  "limma",
  "org.Hs.eg.db",   # for human
  "org.Mm.eg.db"    # for mouse
))
```

---

# Quick Start

## Minimal example: ORA only

This example requires only a list of significant DEG symbols and an API key.

```{r quickstart-ora}
library(LLM_EnrichViz)

# Write your API key to a file (one line, no quotes)
writeLines(Sys.getenv("ANTHROPIC_API_KEY"), "~/.claude_key")

LLM_EnrichViz(
  deg_list            = c("TP53", "MYC", "EGFR", "CDKN1A", "BCL2",
                          "CCND1", "MDM2", "RB1", "E2F1", "CDK4"),
  gene_type           = "SYMBOL",
  organism            = "human",
  analyses            = "ORA",
  databases           = c("KEGG", "GO"),
  llm_provider        = "claude",
  llm_model           = "claude-sonnet-4-6",
  api_key_file        = "~/.claude_key",
  experimental_design = "Colon cancer vs normal tissue, bulk RNA-Seq, n=30",
  output_dir          = "results_ora_only"
)
```

Results are written to `results_ora_only/`:

```
results_ora_only/
├── run_parameters.txt
├── ORA_KEGG_results.txt
├── ORA_GO_BP_results.txt
├── ORA_KEGG_barplot.pdf
├── ORA_GO_BP_barplot.pdf
├── agent_responses.txt
├── final_response.txt
├── integrated_network.html
└── report_template.html
```

---

# Full Pipeline

## Preparing input data

### From DESeq2

```{r deseq2-input}
library(DESeq2)

# Assuming 'dds' is your DESeqDataSet
res       <- DESeq2::results(dds)
res_df    <- as.data.frame(res)
res_df    <- res_df[!is.na(res_df$padj), ]

# ORA: significant DEGs
deg_list  <- rownames(res_df[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1, ])

# GSEA: all genes ranked by Wald statistic (signed, decreasing)
ranked_list <- res_df$stat
names(ranked_list) <- rownames(res_df)
ranked_list <- sort(ranked_list[!is.na(ranked_list)], decreasing = TRUE)

# GSVA / ROAST: normalised count matrix
expr_matrix <- DESeq2::vst(dds, blind = FALSE) |> SummarizedExperiment::assay()

# Design matrix for ROAST
design_mat  <- model.matrix(~ condition, data = colData(dds))
contrast_v  <- c(0, 1)  # test the second coefficient
```

### From Seurat (single-cell)

```{r seurat-input}
library(Seurat)

# After FindMarkers()
markers <- FindMarkers(seurat_obj, ident.1 = "Tumour", ident.2 = "Normal")

# ORA
deg_list    <- rownames(markers[markers$p_val_adj < 0.05 &
                                  abs(markers$avg_log2FC) > 0.5, ])

# GSEA (rank by avg_log2FC * -log10(p_val))
markers$rank_stat <- markers$avg_log2FC * (-log10(markers$p_val + 1e-300))
ranked_list       <- sort(setNames(markers$rank_stat, rownames(markers)),
                          decreasing = TRUE)

# Pseudo-bulk expression for GSVA / ROAST
expr_matrix <- AggregateExpression(seurat_obj, group.by = "sample_id",
                                    assays = "RNA")$RNA
```

## Running the full pipeline

```{r full-pipeline}
LLM_EnrichViz(
  # Inputs
  expression_data     = expr_matrix,
  ranked_list         = ranked_list,
  deg_list            = deg_list,
  gene_type           = "SYMBOL",
  organism            = "human",

  # ROAST design
  design_matrix       = design_mat,
  contrast_vector     = contrast_v,

  # Methods and databases
  analyses            = c("ORA", "GSEA", "GSVA", "ROAST"),
  databases           = c("KEGG", "GO", "Reactome", "WikiPathways"),

  # LLM settings
  llm_provider        = "openai",
  llm_model           = "gpt-4o",
  api_key_file        = "~/.openai_key",
  temperature         = 0,

  # Experiment context
  experimental_design = paste(
    "PBMC from 20 COVID-19 patients (WHO severity 3–5) vs 20 healthy",
    "donors; paired-end bulk RNA-Seq; 50 M reads/sample."
  ),

  output_dir          = "results_covid_pbmc"
)
```

---

# Switching LLM Providers

The `llm_provider` and `llm_model` arguments control which LLM is used for
**all** agents and the final synthesis. You can switch without changing
anything else.

```{r providers}
# Google Gemini
LLM_EnrichViz(
  ...,
  llm_provider = "gemini",
  llm_model    = "gemini-1.5-pro-latest",
  api_key_file = "~/.gemini_key"
)

# OpenAI GPT-4o
LLM_EnrichViz(
  ...,
  llm_provider = "openai",
  llm_model    = "gpt-4o",
  api_key_file = "~/.openai_key"
)

# Anthropic Claude
LLM_EnrichViz(
  ...,
  llm_provider = "claude",
  llm_model    = "claude-sonnet-4-6",
  api_key_file = "~/.anthropic_key"
)

# DeepSeek
LLM_EnrichViz(
  ...,
  llm_provider = "deepseek",
  llm_model    = "deepseek-chat",
  api_key_file = "~/.deepseek_key"
)
```

---

# Understanding the Output

## Agent responses (`agent_responses.txt`)

Each agent corresponds to one (method × database) combination. For a run with
ORA + GSEA on KEGG + GO, there are 4 agents. Responses include:

- Summary and categorisation of enriched terms
- Hub gene identification (step-by-step reasoning)
- Method-specific insights (NES for GSEA, direction for ROAST, etc.)
- A TSV system-representation network

## Final synthesis (`final_response.txt`)

The orchestrator LLM produces:

1. **Cross-method convergence table** — pathways ranked by how many methods detected them
2. **Complementary insights** — what each method contributed uniquely
3. **Multi-method hub genes** — ranked by number of supporting methods
4. **Integrated mechanistic model**
5. **Confidence-annotated network** (TSV with Confidence_Level 1–4)
6. **10-word title** including the top hub gene

## Interactive network (`integrated_network.html`)

Nodes coloured by type (Gene = blue, Process/pathway = orange). Click any node
to highlight its neighbours. Filter by group using the dropdown.

---

# Advanced Usage

## Running individual analyses

Each enrichment function can be called independently:

```{r advanced-individual}
gene_mapping <- map_gene_ids_enrichviz(
  deg_list  = deg_list,
  gene_type = "SYMBOL",
  organism  = "human"
)

# ORA only
ora_results <- run_ORA(
  gene_ids     = gene_mapping$ENTREZID,
  gene_mapping = gene_mapping,
  organism     = "human",
  databases    = c("KEGG", "Reactome"),
  pvalue       = 0.01,
  output_dir   = "my_results"
)

# GSEA only
gsea_results <- run_GSEA(
  ranked_list  = ranked_list,
  gene_mapping = gene_mapping,
  gene_type    = "SYMBOL",
  organism     = "human",
  databases    = c("GO", "WikiPathways"),
  pvalue       = 0.05,
  output_dir   = "my_results"
)
```

## Running LLM helpers per analysis

Use the short helpers to run the LLM loop on a single method's results:

```{r advanced-llm-helpers}
llm_gsea <- LLM_GSEA(
  enrichment_results = gsea_results,
  gene_symbols       = unique(gene_mapping$SYMBOL),
  llm_provider       = "openai",
  api_key_file       = "~/.openai_key",
  output_dir         = "my_results"
)
```

## Using individual LLM functions directly

```{r advanced-llm}
prompt <- "Summarise the role of TP53 in apoptosis in 3 sentences."

# Claude
response <- make_claude_request(
  prompt            = prompt,
  temperature       = 0,
  max_output_tokens = 512,
  api_key           = readLines("~/.anthropic_key"),
  llm_model         = "claude-sonnet-4-6",
  delay_seconds     = 2
)
cat(response)
```

## Retrieving gene sets

```{r advanced-genesets}
gene_sets <- get_gene_sets(
  databases = c("KEGG", "Reactome"),
  organism  = "human"
)
# gene_sets$KEGG is a named list of character vectors
length(gene_sets$KEGG)
head(names(gene_sets$KEGG))
```

---

# Reproducibility Notes

- Set `temperature = 0` for deterministic LLM responses.
- Save `run_parameters.txt` (written automatically) for exact reproducibility.
- LLM responses may vary across model versions; pin `llm_model` explicitly.
- For ROAST, set a fixed seed before running: `set.seed(42)`.

---

# Session Info

```{r session, eval=TRUE}
sessionInfo()
```
