---
title: "LLM_EnrichViz — Enrichment Analysis Report"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    self_contained: true
params:
  output_dir: "enrichviz_results"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center")
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

od <- params$output_dir
```

# Overview

This report summarises the multi-method gene set enrichment analysis
performed by **LLM_EnrichViz**.  Each section presents the results of one
method (ORA, GSEA, GSVA, ROAST) across the requested pathway databases,
followed by the integrated LLM interpretation.

---

# Analysis Parameters

```{r params-table}
params_file <- file.path(od, "analysis_parameters.txt")
if (file.exists(params_file)) {
  param_lines <- readLines(params_file)
  param_df    <- data.frame(Parameter = param_lines)
  kable(param_df, caption = "Analysis parameters") %>%
    kable_styling(bootstrap_options = c("striped","hover","condensed"),
                  full_width = FALSE)
} else {
  cat("Parameter file not found.")
}
```

---

# ORA Results

```{r ora-results}
ora_files <- list.files(od, pattern = "^ORA_.*_results\\.txt$", full.names = TRUE)
if (length(ora_files) > 0) {
  for (f in ora_files) {
    db_name <- gsub("ORA_|_results\\.txt", "", basename(f))
    df      <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                          fill = TRUE, quote = "")
    if (nrow(df) == 0) next
    cat("##", db_name, "\n\n")
    df_show <- head(df, 20)
    print(kable(df_show, caption = paste("ORA —", db_name, "(top 20)")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"),
                    full_width = TRUE, font_size = 11))
    cat("\n\n")
  }
} else {
  cat("No ORA result files found.\n")
}
```

```{r ora-plots, fig.width=10, fig.height=6}
ora_plots <- list.files(od, pattern = "^ORA_.*_barplot\\.pdf$", full.names = TRUE)
if (length(ora_plots) > 0) {
  for (f in ora_plots) {
    db_name <- gsub("ORA_|_barplot\\.pdf", "", basename(f))
    cat("### Bar chart —", db_name, "\n\n")
    knitr::include_graphics(f)
    cat("\n\n")
  }
}
```

---

# GSEA Results

```{r gsea-results}
gsea_files <- list.files(od, pattern = "^GSEA_.*_results\\.txt$", full.names = TRUE)
if (length(gsea_files) > 0) {
  for (f in gsea_files) {
    db_name <- gsub("GSEA_|_results\\.txt", "", basename(f))
    df      <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                          fill = TRUE, quote = "")
    if (nrow(df) == 0) next
    cat("##", db_name, "\n\n")
    df_show <- head(df, 20)
    print(kable(df_show, caption = paste("GSEA —", db_name, "(top 20)")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"),
                    full_width = TRUE, font_size = 11))
    cat("\n\n")
  }
} else {
  cat("No GSEA result files found.\n")
}
```

---

# GSVA Results

```{r gsva-results}
gsva_files <- list.files(od, pattern = "^GSVA_.*_summary\\.txt$", full.names = TRUE)
if (length(gsva_files) > 0) {
  for (f in gsva_files) {
    db_name <- gsub("GSVA_|_summary\\.txt", "", basename(f))
    df      <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                          fill = TRUE, quote = "")
    if (nrow(df) == 0) next
    cat("##", db_name, "\n\n")
    df_show <- head(df, 20)
    print(kable(df_show, caption = paste("GSVA —", db_name, "(top 20 |MeanScore|)")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"),
                    full_width = TRUE, font_size = 11))
    cat("\n\n")
  }
} else {
  cat("No GSVA result files found.\n")
}
```

---

# ROAST Results

```{r roast-results}
roast_files <- list.files(od, pattern = "^ROAST_.*_results\\.txt$", full.names = TRUE)
if (length(roast_files) > 0) {
  for (f in roast_files) {
    db_name <- gsub("ROAST_|_results\\.txt", "", basename(f))
    df      <- read.table(f, header = TRUE, sep = "\t", stringsAsFactors = FALSE,
                          fill = TRUE, quote = "")
    if (nrow(df) == 0) next
    cat("##", db_name, "\n\n")
    df_show <- head(df, 20)
    print(kable(df_show, caption = paste("ROAST —", db_name, "(top 20)")) %>%
      kable_styling(bootstrap_options = c("striped","hover","condensed"),
                    full_width = TRUE, font_size = 11))
    cat("\n\n")
  }
} else {
  cat("No ROAST result files found.\n")
}
```

---

# Agent LLM Responses

```{r agent-responses}
agent_file <- file.path(od, "agent_responses.txt")
if (file.exists(agent_file)) {
  cat(readLines(agent_file), sep = "\n")
} else {
  cat("Agent response file not found.\n")
}
```

---

# Integrated LLM Analysis

```{r final-response}
final_file <- file.path(od, "final_response.txt")
if (file.exists(final_file)) {
  cat(readLines(final_file), sep = "\n")
} else {
  cat("Final response file not found.\n")
}
```

---

# Integrated Network

```{r network, results='asis'}
net_file <- file.path(od, "integrated_network.html")
if (file.exists(net_file)) {
  cat('<iframe src="integrated_network.html" width="100%" height="720px"',
      'style="border:none;"></iframe>\n')
} else {
  cat("Network file not found.\n")
}
```

---

# Session Information

```{r session-info}
sessionInfo()
```
